---
title: "Analysis of COVID-19 Deaths and Air Quality in California, 2020"
author: "Jillian Allison"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

notes: 
- get socioeconomic data ---- so we can control for the effects of that and see if air quality without SE disadvantage makes a difference in COVID deaths 
Analyses 
- 5 year average air quality scc ----> cumulative COVID deaths per 100,000 by zip code in 2020
    - then analyze w income effects
- 5 year average air quality CA ----> cumulative COVID deaths per 100,000 by county in 2020 
    - then analyze w income effects 


```{r}
# Packages
library(tidyverse)
library(readr)
library(gt)
library(tufte)
library(janitor)

# File paths
rootdir <- ("/Users/jillianallison/Desktop/MEDS/Courses/EDS222")
datadir <- file.path(rootdir,"data","final")

lct <- Sys.getlocale("LC_TIME"); Sys.setlocale("LC_TIME", "C")

```

### Load & Clean Data

#### Air Quality Data
notes: 
clean statewide data: 
* combine dataframes 
* select date, Daily mean pm 2.5 concentration, units, daily aqi, site name, county code, county, latitude, longitude 
* group by county, summarize mean pm 2.5, summarize mean daily aqi 
```{r}
# CA
# Load raw data
raw_aq_ca_2015 <- read.csv(file.path(datadir, "air_quality","aq_ca_2015.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2016 <- read.csv(file.path(datadir, "air_quality","aq_ca_2016.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2017 <- read.csv(file.path(datadir, "air_quality","aq_ca_2017.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2018 <- read.csv(file.path(datadir, "air_quality","aq_ca_2018.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2019 <- read.csv(file.path(datadir, "air_quality","aq_ca_2019.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2020 <- read.csv(file.path(datadir, "air_quality","aq_ca_2020.csv"), stringsAsFactors = FALSE)

# Combine data frames 
raw_aq_ca <- rbind(raw_aq_ca_2015, raw_aq_ca_2016, raw_aq_ca_2017,
                   raw_aq_ca_2018, raw_aq_ca_2019, raw_aq_ca_2020) %>%
  clean_names()

raw_aq_ca$date <- as.Date(raw_aq_ca$date, format = "%m/%d/%Y")
  

# Select necessary columns 
aqi_ca <- raw_aq_ca %>%
  select(1, 7, 17:20)

pm_ca <- raw_aq_ca %>%
  select(1, 5, 6, 17:20)
```

```{r}
# Santa Clara County 
raw_aq_scc_2015 <- read.csv(file.path(datadir, "air_quality","aq_scc_2015.csv"), stringsAsFactors = FALSE)
raw_aq_scc_2016 <- read.csv(file.path(datadir, "air_quality","aq_scc_2016.csv"), stringsAsFactors = FALSE)
raw_aq_scc_2017 <- read.csv(file.path(datadir, "air_quality","aq_scc_2017.csv"), stringsAsFactors = FALSE)
raw_aq_scc_2018 <- read.csv(file.path(datadir, "air_quality","aq_scc_2018.csv"), stringsAsFactors = FALSE)
raw_aq_scc_2019 <- read.csv(file.path(datadir, "air_quality","aq_scc_2019.csv"), stringsAsFactors = FALSE)
raw_aq_scc_2020 <- read.csv(file.path(datadir, "air_quality","aq_scc_2020.csv"), stringsAsFactors = FALSE)

# Combine data frames 
raw_aq_scc <- rbind(raw_aq_scc_2015, raw_aq_scc_2016, raw_aq_scc_2017,
                   raw_aq_scc_2018, raw_aq_scc_2019, raw_aq_scc_2020) %>%
  clean_names()

raw_aq_scc$date <- as.Date(raw_aq_scc$date, format = "%m/%d/%Y")
  

# Select necessary columns 
aqi_scc <- raw_aq_scc %>%
  select(1, 7:8, 19:20)

pm_scc <- raw_aq_scc %>%
  select(1, 5:6, 8, 19:20)
```
#### COVID Data 
```{r}
# CALIFORNIA
raw_ca_covid_2020 <- read_csv(file.path(datadir, "ca_cases_tests_deaths.csv"))

# Filter for only counties represented in AQ data, select relevant columns 
ca_covid_2020 <- raw_ca_covid_2020 %>%
  filter(area %in% pm_ca$county) %>%
  select(1, 2, 4:8)
```


```{r}
# SANTA CLARA COUNTY
# Santa Clara County cumulative COVID data by zip code for 2020
raw_scc_cum_covid_2020 <- read_csv(file.path(datadir, "cum_deaths_zipcode_scc.csv"))

# Zip codes lat and lon info 
zip_codes <- read_csv(file.path(datadir, "zip_lat_long.csv")) %>%
  rename(zip_code = ZIP) %>% 
  mutate(zip_code = as.numeric(zip_code)) %>%
  filter(zip_code %in% scc_cum_covid_2020$zip_code) 

# Data wrangling
scc_cum_covid_2020 <- raw_scc_cum_covid_2020 %>%
  rename(zip_code = zcta) 

scc_cum_covid_spatial <- inner_join(scc_cum_covid_2020, zip_codes, by = "zip_code")
```

