---
title: "Analysis of COVID-19 Deaths and Air Quality in California, 2020"
author: "Jillian Allison"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

notes: 
- get socioeconomic data ---- so we can control for the effects of that and see if air quality without SE disadvantage makes a difference in COVID deaths 
Analyses 
- 5 year average air quality scc ----> cumulative COVID deaths per 100,000 by zip code in 2020
    - then analyze w income effects
- 5 year average air quality CA ----> cumulative COVID deaths per 100,000 by county in 2020 
    - then analyze w income effects 


```{r}
# Packages
library(tidyverse)
library(readr)
library(gt)
library(tufte)
library(janitor)

# File paths
rootdir <- ("/Users/jillianallison/Desktop/MEDS/Courses/EDS222")
datadir <- file.path(rootdir,"data","final")

lct <- Sys.getlocale("LC_TIME"); Sys.setlocale("LC_TIME", "C")

```

### Load & Clean Data

#### Air Quality Data
```{r}
# CA
# Load raw data
raw_aq_ca_2015 <- read.csv(file.path(datadir, "air_quality","aq_ca_2015.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2016 <- read.csv(file.path(datadir, "air_quality","aq_ca_2016.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2017 <- read.csv(file.path(datadir, "air_quality","aq_ca_2017.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2018 <- read.csv(file.path(datadir, "air_quality","aq_ca_2018.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2019 <- read.csv(file.path(datadir, "air_quality","aq_ca_2019.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2020 <- read.csv(file.path(datadir, "air_quality","aq_ca_2020.csv"), stringsAsFactors = FALSE)

# Combine data frames 
raw_aq_ca <- rbind(raw_aq_ca_2015, raw_aq_ca_2016, raw_aq_ca_2017,
                   raw_aq_ca_2018, raw_aq_ca_2019, raw_aq_ca_2020) %>%
  clean_names()

raw_aq_ca$date <- as.Date(raw_aq_ca$date, format = "%m/%d/%Y")
  

# Select necessary columns 
aqi_ca <- raw_aq_ca %>%
  select(1, 7, 17:20)

pm_ca <- raw_aq_ca %>%
  select(1, 5, 6, 17:20)

# Create the 6 year averages by county --- 2015 - 2020 
ltavg_pm_ca <- pm_ca %>%
  group_by(county) %>%
  summarize(average_pm = mean(daily_mean_pm2_5_concentration))

ltavg_aqi_ca <- aqi_ca %>%
  group_by(county) %>%
  summarize(average_aqi = mean(daily_aqi_value))
```

```{r}
# Santa Clara County 
raw_aq_scc_2015 <- read.csv(file.path(datadir, "air_quality","aq_scc_2015.csv"), stringsAsFactors = FALSE)
raw_aq_scc_2016 <- read.csv(file.path(datadir, "air_quality","aq_scc_2016.csv"), stringsAsFactors = FALSE)
raw_aq_scc_2017 <- read.csv(file.path(datadir, "air_quality","aq_scc_2017.csv"), stringsAsFactors = FALSE)
raw_aq_scc_2018 <- read.csv(file.path(datadir, "air_quality","aq_scc_2018.csv"), stringsAsFactors = FALSE)
raw_aq_scc_2019 <- read.csv(file.path(datadir, "air_quality","aq_scc_2019.csv"), stringsAsFactors = FALSE)
raw_aq_scc_2020 <- read.csv(file.path(datadir, "air_quality","aq_scc_2020.csv"), stringsAsFactors = FALSE)

# Combine data frames 
raw_aq_scc <- rbind(raw_aq_scc_2015, raw_aq_scc_2016, raw_aq_scc_2017,
                   raw_aq_scc_2018, raw_aq_scc_2019, raw_aq_scc_2020) %>%
  clean_names()

raw_aq_scc$date <- as.Date(raw_aq_scc$date, format = "%m/%d/%Y")
  

# Select necessary columns 
aqi_scc <- raw_aq_scc %>%
  select(1, 7:8, 19:20)

pm_scc <- raw_aq_scc %>%
  select(1, 5:6, 8, 19:20)

# Create the 6 year averages by county --- 2015 - 2020 
ltavg_pm_scc <- pm_scc %>%
  group_by(site_name, site_latitude, site_longitude) %>%
  summarize(average_pm = mean(daily_mean_pm2_5_concentration))

ltavg_aqi_scc <- aqi_scc %>%
  group_by(site_name, site_latitude, site_longitude) %>%
  summarize(average_aqi = mean(daily_aqi_value))
```
#### COVID Data 
```{r}
# CALIFORNIA
raw_ca_covid_2020 <- read_csv(file.path(datadir, "ca_cases_tests_deaths.csv"))

# Filter for only counties represented in AQ data, select relevant columns 
ca_covid_2020 <- raw_ca_covid_2020 %>%
  filter(area %in% pm_ca$county) %>%
  select(1, 2, 4:8)
```


```{r}
# SANTA CLARA COUNTY
# Santa Clara County cumulative COVID data by zip code for 2020
raw_scc_cum_covid_2020 <- read_csv(file.path(datadir, "cum_deaths_zipcode_scc.csv"))

# Data wrangling
scc_cum_covid_2020 <- raw_scc_cum_covid_2020 %>%
  rename(zip_code = zcta) 

# Zip codes contains lat and lon info 
zip_codes <- read_csv(file.path(datadir, "zip_lat_long.csv")) %>%
 rename(zip_code = ZIP) %>%
 mutate(zip_code = as.numeric(zip_code)) %>%
 filter(zip_code %in% scc_cum_covid_2020$zip_code) 

scc_cum_covid_spatial <- inner_join(scc_cum_covid_2020, zip_codes, by = "zip_code")
```


### Spatially representing SCC data 
I'll compare north county and south county COVID death rates since the air quality is different --- better in Gilroy than in the valley
```{r}
# Santa Clara County--- 
# Making things SPATIAL 

# Covid cumulative deaths 
library(terra)
library(tidyterra)
library(ggplot2)

zip_codes_vect <- vect(scc_cum_covid_spatial, geom=c("LNG", "LAT"), crs="EPSG:4326", keepgeom=FALSE)


ltavg_pm_scc_vect <- vect(ltavg_pm_scc, geom = c("site_longitude", "site_latitude"), crs = "EPSG:4326",  keepgeom = FALSE)

ltavg_aqi_scc_vect <- vect(ltavg_aqi_scc, geom = c("site_longitude", "site_latitude"), crs = "EPSG:4326",  keepgeom = FALSE)

ggplot() + 
  geom_spatraster(data = map_rcl) + 
  scale_fill_gradient(low = "#cfe2f3", 
                      high = "#6aa84f") + 
  geom_spatvector(data = ltavg_aqi_scc_vect)

```

```{r}
# Beginnings of a map but you should fix the color and the fact that its floating ok 

county_boundary <- file.path("/Users/jillianallison/Desktop/MEDS/Courses/EDS222/data/final/scc_county_boundary/geo_export_68da2c43-7af7-4d7a-b35c-05339ab65394.shp")
county_boundary_v <- vect(f)

spatial_extent <- rast(bb, crs = "EPSG:3857")
map <- set_defaults(map_service = "esri", map_type = "natgeo_world_map")

basemap_tif <- basemap_geotif(ext = county_boundary_v) # downloads a basemap as a .tif file

map <- rast(file.path(basemap_tif)) 

rcl <- matrix(c(200, Inf, 1), ncol = 3, byrow = TRUE)

map_rcl <- classify(map, rcl = rcl)


ggplot() + 
  geom_spatraster(data = map_rcl) + 
  scale_fill_gradient(low = "#cfe2f3", 
                      high = "#6aa84f") +
  geom_spatvector(data = zip_codes_vect) 
```

