---
title: "Analysis of COVID-19 Deaths and Air Quality in California, 2020"
author: "Jillian Allison"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

notes: - get socioeconomic data ---- so we can control for the effects
of that and see if air quality without SE disadvantage makes a
difference in COVID deaths Analyses - 5 year average air quality scc
----\> cumulative COVID deaths per 100,000 by zip code in 2020 - then
analyze w income effects - 5 year average air quality CA ----\>
cumulative COVID deaths per 100,000 by county in 2020 - then analyze w
income effects

```{r}
# Packages
library(tidyverse)
library(readr)
library(gt)
library(tufte)
library(janitor)

# File paths
rootdir <- ("/Users/jillianallison/Desktop/MEDS/Courses/EDS222")
datadir <- file.path(rootdir,"data","final")

lct <- Sys.getlocale("LC_TIME"); Sys.setlocale("LC_TIME", "C")

```

### Load & Clean Data

#### Air Quality Data

```{r}
# CA
# Load raw data
raw_aq_ca_2015 <- read.csv(file.path(datadir, "air_quality","aq_ca_2015.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2016 <- read.csv(file.path(datadir, "air_quality","aq_ca_2016.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2017 <- read.csv(file.path(datadir, "air_quality","aq_ca_2017.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2018 <- read.csv(file.path(datadir, "air_quality","aq_ca_2018.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2019 <- read.csv(file.path(datadir, "air_quality","aq_ca_2019.csv"), stringsAsFactors = FALSE)
raw_aq_ca_2020 <- read.csv(file.path(datadir, "air_quality","aq_ca_2020.csv"), stringsAsFactors = FALSE)

# Combine data frames 
raw_aq_ca <- rbind(raw_aq_ca_2015, raw_aq_ca_2016, raw_aq_ca_2017,
                   raw_aq_ca_2018, raw_aq_ca_2019, raw_aq_ca_2020) %>%
  clean_names()

raw_aq_ca$date <- as.Date(raw_aq_ca$date, format = "%m/%d/%Y")
  

# Select necessary columns 
aqi_ca <- raw_aq_ca %>%
  select(1, 7, 17:20)

pm_ca <- raw_aq_ca %>%
  select(1, 5, 6, 17:20)

# Create the 6 year averages by county --- 2015 - 2020 
ltavg_pm_ca <- pm_ca %>%
  group_by(county) %>%
  summarize(average_pm = mean(daily_mean_pm2_5_concentration))

ltavg_aqi_ca <- aqi_ca %>%
  group_by(county) %>%
  summarize(average_aqi = mean(daily_aqi_value))
```

#### COVID Data

We're using 2020 data because we can assume that people have not been
vaccinated against COVID.

```{r}
# CALIFORNIA
raw_ca_covid_2020 <- read_csv(file.path(datadir, "ca_cases_tests_deaths.csv"))

# Filter for only counties represented in AQ data, select relevant columns 
ca_covid_2020 <- raw_ca_covid_2020 %>%
  filter(area %in% pm_ca$county) %>%
  select(1, 2, 4:8) %>%
  filter(between(date, as.Date("2020-01-01"), as.Date("2020-12-31")))

# Make summary county data for 2020
ca_counties_covid <- ca_covid_2020 %>%
  group_by(area, population) %>%
  summarize(cum_cases = sum(cases), 
            cum_deaths = sum(deaths)) %>%
  rename(county = area) %>%
  mutate(cum_death_rate = ((cum_deaths / population) * 100000)) %>% # We need to use case rates because areas with higher pops will obv have higher deaths. 
  mutate(cum_case_rate = ((cum_cases / population) * 100000)) %>% # Same logic as above ^^^ 
  mutate(deaths_pct_cases = ((cum_deaths / cum_cases) * 100)) # but I think this is what we should use. bc obviously deaths are dependent on cases .... 
```

## Data Analysis

-   5 year average air quality CA ----\> cumulative COVID deaths per
    100,000 by county in 2020
    -   then analyze w income effects, age effects (find % over 65 data
        by county, should be easy)

$$Deaths_i =\beta_{0}+\beta_{1} \cdot AQI_i+\varepsilon_i$$

```{r}
# Relationship between 6 year avg air quality and cumulative COVID deaths per 100,000 by county in 2020
# These are the 6 year averages for AQI in each county
ltavg_aqi_ca
# These are the 6 year averages for PM 2.5 in each county 
ltavg_pm_ca 
# These are the COVID data 
ca_counties_covid

# Create a df for AQI 
california_aqi_covid <- inner_join(ltavg_aqi_ca, ca_counties_covid,  by = "county")
# Create a df for PM 2.5 
california_pm_covid <- inner_join(ltavg_pm_ca, ca_counties_covid,  by = "county")
```

### Basic scatterplots

```{r}
library(ggrepel)
# Lets just make a simple plot--- independent: AQI, dependent: deaths as a percent of cases 
death_aqi_plt <- ggplot(data = california_aqi_covid, aes(x = average_aqi, y = deaths_pct_cases)) + 
  geom_point() + 
  labs(title = "Deaths & AQI by CA County, 2020", 
       x = "Average Daily AQI from 2015 - 2020", 
       y = "Cumulative deaths as a percent of cases (%)") + 
  theme_minimal()

death_aqi_plt + 
  geom_label_repel(aes(label = county)) + 
  geom_smooth(method='lm', formula= y~x, color="lightcoral", se=F, size=1.5)

# What if we remove the outliers (Inyo and Alpine)
rmoutliers_aqi <- california_aqi_covid %>%
  subset(!county %in% c("Alpine", "Inyo"))

rmoutliers_aqi_plt <- ggplot(data = rmoutliers_aqi, aes(x = average_aqi, y = deaths_pct_cases)) + 
  geom_point() + 
  labs(title = "Deaths & AQI by CA County, 2020", 
       x = "Average Daily AQI from 2015 - 2020", 
       y = "Cumulative deaths as a percent of cases (%)") + 
  theme_minimal()

rmoutliers_aqi_plt + 
  geom_label_repel(aes(label = county)) +
  geom_smooth(method='lm', formula= y~x, color="lightcoral", se=F, size=1.5)
```

```{r}
# Lets just make a simple plot--- independent: PM 2.5, dependent: deaths as a percent of cases 
death_pm_plt <- ggplot(data = california_pm_covid, aes(x = average_pm, y = deaths_pct_cases)) + 
  geom_point() + 
  labs(title = "Deaths & PM 2.5 by CA County, 2020", 
       x = "Average Daily PM 2.5 from 2015 - 2020", 
       y = "Cumulative deaths as a percent of cases (%)") + 
  theme_minimal()

death_pm_plt + 
  geom_label_repel(aes(label = county)) +
  geom_smooth(method='lm', formula= y~x, color="lightcoral", se=F, size=1.5)

# What if we remove the outliers (Inyo and Alpine)
rmoutliers_pm <- california_pm_covid %>%
  subset(!county %in% c("Alpine", "Inyo"))

rmoutliers_pm_plt <- ggplot(data = rmoutliers_pm, aes(x = average_pm, y = deaths_pct_cases)) + 
  geom_point() + 
  labs(title = "Deaths & PM 2.5 by CA County, 2020 (outliers removed)", 
       x = "Average Daily PM 2.5 from 2015 - 2020", 
       y = "Cumulative deaths as a percent of cases (%)") + 
  theme_minimal()

rmoutliers_pm_plt + 
  geom_label_repel(aes(label = county)) +
  geom_smooth(method='lm', formula= y~x, color="lightcoral", se=F, size=1.5)
```

### Linear regressions- single variable

```{r}
# AQI 
# With Alpine and Inyo Counties 
summary(lm(deaths_pct_cases ~ average_aqi, data =california_aqi_covid))

# Without Alpine and Inyo Counties 
summary(lm(deaths_pct_cases ~ average_aqi, data =rmoutliers_aqi))

model_aqi_rmoutliers <- lm(deaths_pct_cases ~ average_aqi, data =rmoutliers_aqi)

# Confidence Interval
CI_aqi <- confint(model_aqi_rmoutliers, level = 0.95)

apply(CI_aqi, 1, function(x) paste0("95% CI: [", 
        round(x[1], 2), ",", round(x[2], 2), "]"))
```

**INTERPRET RESULTS FROM LINEAR REGRESSION**

```{r}
# PM 2.5 
# With Alpine and Inyo Counties 
summary(lm(deaths_pct_cases ~ average_pm, data =california_pm_covid))

# Without Alpine and Inyo Counties 
summary(lm(deaths_pct_cases ~ average_pm, data =rmoutliers_pm))

# Confidence Interval
model_pm_rmoutliers <- lm(deaths_pct_cases ~ average_pm, data =rmoutliers_pm)

tab_model(model_pm_rmoutliers)

CI_pm <- confint(model_pm_rmoutliers, level = 0.95)

apply(CI_pm, 1, function(x) paste0("95% CI: [", 
        round(x[1], 2), ",", round(x[2], 2), "]"))
```

**INTERPRET RESULTS FROM LINEAR REGRESSION**

## But we know that different factors can affect COVID deaths...

I'll add two well-known determinants of COVID outcomes into my analysis-
% of population in county over the age of 65, and income data

$$Deaths_i=\beta_{0}+\beta_{1} \cdot Poverty_i + \beta_{2} \cdot AQI_i + \beta_{3} \cdot Poverty_i \cdot AQI_i + \varepsilon_i$$

### Read in poverty and demographic data

```{r}
# Read in poverty data 
raw_ca_counties_poverty <- read_csv(file.path(datadir, "pov_data_county_allrows.csv"))

# Read in demographic data 
raw_ca_counties_demographics <- read_csv(file.path(datadir, "dem_data_county_allrows.csv"))
```

### Clean poverty data

```{r}
# Clean poverty data 
ca_counties_poverty <- raw_ca_counties_poverty %>%
  clean_names() %>%
  filter(label_grouping == "Population for whom poverty status is determined") %>% 
  subset(select = -c(label_grouping)) %>%
  pivot_longer(cols = 1:ncol(.), 
               names_to = "county", 
               values_to = "poverty_level_pct")

# Let's fix the county column values and make the percentages numeric 
ca_counties_poverty <- ca_counties_poverty %>%
  mutate(county = str_remove(.$county, 
                             pattern = "_county_california_percent_below_poverty_level_estimate"))%>%
  mutate(poverty_level_pct = as.numeric(str_remove(.$poverty_level_pct, pattern = "%")))%>%
  mutate(county = str_replace_all(.$county, 
                              pattern = "[_]", " "))%>%
  mutate(county = str_to_title(.$county)) %>% 
  filter(county %in% ca_counties_covid$county)  # filter for only available counties

```

### Clean demographic data

```{r}
med_age_matrix <- raw_ca_counties_demographics %>%
  clean_names() %>%
  slice(19) %>% 
  subset(select = -c(label_grouping)) %>%
  t() %>%
  cbind(county = rownames(.))
rownames(med_age_matrix) <- 1:nrow(med_age_matrix)

ca_counties_med_age <- as.data.frame(med_age_matrix) %>%
  select(county, med_age = V1) %>%
  filter(str_detect(county, "estimate")) %>% 
  mutate(county = str_remove(.$county, 
                             pattern = "_county_california_estimate")) %>%
  mutate(med_age = as.numeric(med_age))%>%
  mutate(county = str_replace_all(.$county, 
                              pattern = "[_]", " "))%>%
  mutate(county = str_to_title(.$county)) %>% 
  filter(county %in% ca_counties_covid$county)
```

```{r}
over_65_matrix <- raw_ca_counties_demographics %>%
  clean_names() %>%
  slice(25)  %>% 
  subset(select = -c(label_grouping)) %>%
  t() %>%
  cbind(county = rownames(.))
rownames(over_65_matrix) <- 1:nrow(over_65_matrix)


ca_counties_over_65 <- as.data.frame(over_65_matrix) %>%
  select(county, pct_over_65 = V1) %>%
  filter(str_detect(county, "percent")) %>% 
  mutate(county = str_remove(.$county, 
                             pattern = "_county_california_percent")) %>%
  mutate(pct_over_65 = as.numeric(str_remove(.$pct_over_65, pattern = "%"))) %>%
  mutate(pct_over_65 = as.numeric(pct_over_65)) %>%
  mutate(county = str_replace_all(.$county, 
                              pattern = "[_]", " ")) %>%
  mutate(county = str_to_title(.$county)) %>% 
  filter(county %in% ca_counties_covid$county)
```

### Data Analysis with MORE INFO!

```{r}
# Data: for reference 
# Here's my poverty data: 
ca_counties_poverty

# Here's demographic data: 
ca_counties_med_age
ca_counties_over_65

# Here are the joined data frames 
california_poverty_covid <- inner_join(ca_counties_covid, ca_counties_poverty, by = "county")

california_medage_covid <- inner_join(ca_counties_covid, ca_counties_med_age, by = "county")

california_pct65_covid <- inner_join(ca_counties_covid, ca_counties_over_65, by = "county")
```

#### Basic plots showing relationships between just poverty / demographics and COVID deaths

```{r}
# POVERTY 
death_poverty_plt <- ggplot(data = california_poverty_covid, aes(x = poverty_level_pct, y = deaths_pct_cases)) + 
  geom_point() + 
  labs(title = "Deaths & Poverty by CA County, 2020", 
       x = "Poverty Level (%)", 
       y = "Cumulative deaths as a percent of cases (%)") + 
  theme_minimal()

death_poverty_plt + 
  geom_label_repel(aes(label = county)) + 
  geom_smooth(method='lm', formula= y~x, color="lightcoral", se=F, size=1.5)
```

```{r}
# MEDIAN AGE
death_medage_plt <- ggplot(data = california_medage_covid, aes(x = med_age, y = deaths_pct_cases)) + 
  geom_point() + 
  labs(title = "Deaths & Median Age by CA County, 2020", 
       x = "Median Age", 
       y = "Cumulative deaths as a percent of cases (%)") + 
  theme_minimal()

death_medage_plt + 
  geom_label_repel(aes(label = county)) + 
  geom_smooth(method='lm', formula= y~x, color="lightcoral", se=F, size=1.5)
```

```{r}
# PCT OVER 65
death_pct65_plt <- ggplot(data = california_pct65_covid, aes(x = pct_over_65, y = deaths_pct_cases)) + 
  geom_point() + 
  labs(title = "Deaths & Percent of Population Over 65 by CA County, 2020", 
       x = "Percentage of population older than 65 (%)", 
       y = "Cumulative deaths as a percent of cases (%)") + 
  theme_minimal()

death_pct65_plt + 
  geom_label_repel(aes(label = county)) + 
  geom_smooth(method='lm', formula= y~x, color="lightcoral", se=F, size=1.5)
```

**Based on these plots, age demographics do not appear to have much of
an effect on county level cumulative COVID deaths. Poverty does appear
to have an effect on county level cumulative COVID deaths.**

**Let's run an interaction model to explore the relationship between
cumulative COVID deaths as a percentage of cumulative cases, and poverty
rates by county in 2020.**

### Interaction Model

```{r}
library(sjPlot)
```


```{r}
# First we need a data frame with both poverty and AQ information. 
# PM 2.5, POVERTY, AND COVID 
pm_poverty_covid <- inner_join(california_pm_covid, ca_counties_poverty, by = "county")

# AQI, POVERTY, AND COVID
aqi_poverty_covid <- inner_join(california_aqi_covid, ca_counties_poverty, by = "county")

### WITHOUT OUTLIERS ###
# PM 2.5, POVERTY, AND COVID
pm_poverty_covid_rmoutliers <- inner_join(california_pm_covid, ca_counties_poverty, by = "county") %>%
  subset(!county %in% c("Alpine", "Inyo"))

# AQI, POVERTY, COVID 
aqi_poverty_covid_rmoutliers <- inner_join(california_aqi_covid, ca_counties_poverty, by = "county") %>%
  subset(!county %in% c("Alpine", "Inyo"))
```

```{r}
# PLOT PM 
ggplot(pm_poverty_covid, aes(y = deaths_pct_cases, x = average_pm, color = poverty_level_pct)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  scale_color_gradient(low = "blue", high = "red") + 
  labs(title = "PM 2.5 and COVID Deaths by County in California, 2020", 
       x = "Average Daily PM 2.5 from 2015 - 2020", 
       y = "Cumulative Deaths as a Percent of Cumulative Cases (%)") + 
  theme_minimal()
```

```{r}
# PLOT PM WITHOUT OUTLIERS 
pm_int_plot <- ggplot(pm_poverty_covid_rmoutliers, aes(y = deaths_pct_cases, x = average_pm, color = poverty_level_pct)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  scale_color_gradient(low = "blue", high = "red") + 
  labs(title = "PM 2.5 and COVID Deaths by County in California, 2020 (outliers removed)", 
       x = "Average Daily PM 2.5 from 2015 - 2020", 
       y = "Cumulative Deaths as a Percent of Cumulative Cases (%)", 
       color = "Poverty Level (%)") + 
  theme_minimal()

pm_int_plot

# pm_int_plot + 
#   geom_label_repel(aes(label = county)) +
#   geom_smooth(method='lm', formula= y~x, color="lightcoral", se=F, size=1.5)
```

```{r}
# PM 2.5
summary(lm(deaths_pct_cases ~  average_pm + poverty_level_pct + poverty_level_pct:average_pm, data = pm_poverty_covid))

summary(lm(deaths_pct_cases ~  average_pm + poverty_level_pct + poverty_level_pct:average_pm, data = pm_poverty_covid_rmoutliers))
```


REMOVED OUTLIERS

Coefficients: Estimate Std. Error t value Pr(\>\|t\|) (Intercept)
0.891266 0.839153 1.062 0.294 average_pm -0.001107 0.088463 -0.013 0.990
poverty_level_pct 0.007032 0.057871 0.122 0.904
average_pm:poverty_level_pct 0.001261 0.005801 0.217 0.829

Residual standard error: 0.4185 on 46 degrees of freedom Multiple
R-squared: 0.07067, Adjusted R-squared: 0.01006 F-statistic: 1.166 on 3
and 46 DF, p-value: 0.333
```{r}
intmodel_pm_rmoutliers <- lm(deaths_pct_cases ~  average_pm + poverty_level_pct + poverty_level_pct:average_pm, data = pm_poverty_covid_rmoutliers)

tab_model(intmodel_pm_rmoutliers)

# CONFIDENCE INTERVAL PM 
intCI_pm <- confint(intmodel_pm_rmoutliers, level = 0.95)

apply(intCI_pm, 1, function(x) paste0("95% CI: [", 
        round(x[1], 2), ",", round(x[2], 2), "]"))
```

```{r}
# AQI
intmodel_aqi <- summary(lm(deaths_pct_cases ~  poverty_level_pct + average_aqi + poverty_level_pct:average_aqi, data = aqi_poverty_covid))

summary(lm(deaths_pct_cases ~  poverty_level_pct + average_aqi + poverty_level_pct:average_aqi, data = aqi_poverty_covid_rmoutliers))
```


```{r}
intmodel_aqi_rmoutliers <- lm(deaths_pct_cases ~  poverty_level_pct + average_aqi + poverty_level_pct:average_aqi, data = aqi_poverty_covid_rmoutliers)

intCI_aqi <- confint(intmodel_aqi_rmoutliers, level = 0.95)

apply(intCI_aqi, 1, function(x) paste0("95% CI: [", 
        round(x[1], 2), ",", round(x[2], 2), "]"))

```


REMOVED OUTLIERS

Coefficients: **Estimate Std. Error t value Pr(\>\|t\|)** (Intercept)
0.6652902 0.9112505 0.730 0.469 poverty_level_pct 0.0058699 0.0614820
0.095 0.924 average_aqi 0.0074039 0.0264355 0.280 0.781
poverty_level_pct:average_aqi 0.0003046 0.0017161 0.177 0.860

Residual standard error: 0.4094 on 46 degrees of freedom Multiple
R-squared: 0.111, Adjusted R-squared: 0.05304 F-statistic: 1.915 on 3
and 46 DF, p-value: 0.1404
